Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-03-17 manuscript/manuscript.Rmd
added some text.

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)). To accomplish these two objectives, we advocate the following four measures: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspires researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020]. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

# The ethology and ethnography of stone toolmaking

new toolkit such as BORIS were introduced [@friard2016]

```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram of the Perception-Process-Product conceptual framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```

@@

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping [@whittaker2004] are not novel, however, it has never been formally recognized as a legitimate research method in experimental archaeology.

@@

We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits, preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979].

>>>>

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following four measures as integral components of the Triple P framework: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020]. The controlled experiments conducted on stone artifacts [@li2022], partcularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

# The ethology and ethnography of stone toolmaking experiments

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)). new toolkit such as BORIS were introduced [@friard2016]

```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram demonstrating how to operationalize the Perception-Process-Product conceptual framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```

@@

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology.

@@

We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics. One such example is the the edge angle [@crabtree1977]

Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
